---
title: "Class 11 Lab Session"
author: "Olivia Chu"
format: gfm
---

# Section 1. Proportion of G/G in a Population

Download a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39818860-39938861;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel >

Here, we read this CSV file

```{r}
mxl <- read.csv("MXL data.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

14% of Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G).

Now, let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("GBR data.csv")
```

Find proportion of G|G.

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is  more frequent in GBR population than the MKL population.

Let's now dig into this further.


























