---
title: "Class 17: Mini Project"
author: "Olivia Chu (PID 911)"
format: html
---

The goal of this hands-on mini-project is to examine and compare the COVID-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

# Data Import

```{r}
url <- "https://data.chhs.ca.gov/dataset/ead44d40-fd63-4f9f-950a-3b0111074de8/resource/ec32eece-7474-4488-87f0-6e91cb577458/download/covid19vaccinesbyzipcode_test.csv"

# Import vaccination data
vax <- read.csv(url)

head(vax)
```

> Q1. What column details the total number of people fully vaccinated? 

The column titled "persons_fully_vaccinated," details the total number of people fully vaccinated.

```{r}
# vax$persons_fully_vaccinated
```

> Q2. What column details the Zip code tabulation area? 

The column titled "zip_code_tabulation_area," details the zip code tabulation area.

```{r}
# vax$zip_code_tabulation_area
```

> Q3. What is the earliest date in this dataset? 

```{r}
vax$as_of_date[1]
```

The earliest date in this dataset is 2021-01-05 (01/05/2021).

> Q4. What is the latest date in this dataset? 

```{r}
vax$as_of_date[nrow(vax)]
```

The latest date in this dataset is 2023-02-28 (02/28/2023).

We can use the `skim()` function for a quick overview of this dataset.

```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 

There are 13 numeric columns in this dataset.

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

There are 16525 NA values in the persons_fully_vaccinated column.

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
round( sum(is.na(vax$persons_fully_vaccinated))/nrow(vax) * 100, 2 )
```

8.29% of persons_fully_vaccinated values are missing.

> Q8. [Optional]: Why might this data be missing?

This data could be missing due to privacy reasons, such as people who report to other agencies and not to the state (i.e. government employees, federal employees, military, etc.)

# Working with dates

The lubridate package makes working with dates and times in R much less of a pain. Let's have a first play with this package here.

```{r}
library(lubridate)
```

```{r}
today()
```

We can now magically do math with dates.

```{r}
today() - ymd("2021-01-05")
```

Calculating my age in days for fun:

```{r}
today() - ymd("2000-01-13")
```

Let's treat the whole col.

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

How many days have passed since the first vaccination reported in this dataset?

```{r}
today() - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[nrow(vax)]
```
7 days have passed since the last update of the dataset.

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```

There are 113 unique dates in this dataset.

# Working with ZIP codes

ZIP codes are also rather annoying things to work with as they are numeric but not in the conventional sense of doing math.

Just like dates we have special packages to help us work with ZIP codes.

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

We ca calculate the distance between the centroids of any two ZIP codes in miles using `zip_distance()`.

```{r}
zip_distance('92037','92126')
```

More usefully, we can pull census data about ZIP code areas (including median household income etc.) using `reverse_zipcode()`. For example:

```{r}
reverse_zipcode(c('92037', "92126") )
```

# Focus on the San Diego area

Let’s now focus in on the San Diego County area by restricting ourselves first to vax$county == "San Diego" entries. We have two main choices on how to do this. The first using base R the second using the dplyr package:

```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$county == "San Diego", ]
nrow(sd)
```

It is time to revisit the most awesome **dplyr** package.

```{r}
library(dplyr)
```

```{r}
sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Using dplyr is often more convenient when we are subsetting across multiple criteria - for example, all San Diego county areas with a population of over 10,000.

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)

head(sd.10)
```

> Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```

There are 107 distinct zip codes listed for San Diego County.

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 

```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]
```

The zip code area that has the largest 12 + population in this dataset is 92154.

Using dplyr, select all San Diego “county” entries on “as_of_date” “2023-02-28” and use this for the following questions.

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2023-02-28”? 

```{r}
## sd$as_of_date
sd.today <- filter(sd, as_of_date == "2023-02-28")

mean(sd.today$percent_of_population_fully_vaccinated, na.rm=T) * 100
```

The overall average "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2023-02-28" is 74%.

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2023-02-28”?

```{r}
library(ggplot2)
```

```{r}
ggplot(sd.today,
  aes(x=percent_of_population_fully_vaccinated)) +
  geom_histogram(bins = 20, na.rm=T) +
  labs(title = "Vaccination Rates Across San Diego County", 
       x = "Percent of Population Fully Vaccinated on 2023-02-28",
       y = "Count (ZIP code areas)")
```

# Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot, make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ucplot <- ggplot(ucsd) +
            aes(as_of_date, percent_of_population_fully_vaccinated) +
            geom_point() +
            geom_line() +
            labs(title = "Vaccination rate for La Jolla, CA 92037",
            x="Date",
            y="Percent Vaccinated")

ucplot
```

# Comparing to similar sized areas

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on the most recent date of “2023-02-28”.

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                 as_of_date == "2023-02-28")
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
ave <- mean(vax.36$percent_of_population_fully_vaccinated)
ave
```

```{r}
ucplot + geom_hline(yintercept=ave, color="red", linetype=5)
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-02-28”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

Min = 0.3804
1st Qu. = 0.6457
Median = 0.7181
Mean = 0.7213
3rd Qu. = 0.7907
Max = 1.0000

> Q18. Using ggplot, generate a histogram of this data.

```{r}
ggplot(vax.36,
  aes(x=percent_of_population_fully_vaccinated)) +
  geom_histogram(bins=30, na.rm=T) +
  xlim(c(0, 1)) +
  labs(x = "Percent Vaccinated",
       y = "Count")
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
x <- filter(vax.36, zip_code_tabulation_area %in% c("92109", "92040"))

x
```

The 92109 and 92040 ZIP codes areas are below the average value I calculated for all of these above (69.46% and 55.03% vs average of 72.13% calculated above).

> Q20. Finally, make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="cornflowerblue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only areas with a population over 36,000 are shown") +
  geom_hline(yintercept = ave, linetype=5)
```
















